Epistasis is widespread in the genetic control of transcription in humans

### 2013-05-30 docs/manuscript/nature_letter.tex
there may be evidence for chromosome x chromosome interactions

\subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions, but our analysis shows this is unlikely to be sufficient. The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association (476 out of 549). Only 9 interactions were between SNPs that both had marginal effects and 64 were between SNPs that had no known marginal effects. Additionally, we observed that the largest epistatic variance component for the 529 interactions were proportionately divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms ($p = 0.22$ for divergence from expectation), thus choosing a particular parameterisation \emph{a priori} may not be warranted. This is expected as these traditional epistatic variance components are statistical formulations rather than biologically meaningful patterns.

We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common form of epistasis detected involved a homozygous \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit this masking pattern. Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets. Though the exact mechanism underlying this interaction requires further research, we observed that nine of the 14 \emph{trans} SNPs are located in intronic regions (significantly enriched, $p = 3.11 \times 10^{-9}$) and MBNL1 is involved in RNA modification and regulation of splicing, suggesting a putative mechanism by which epistatic interactions may regulate transcription.

\subsection{Contribution of epistatic variance}
Though we have discovered many instances of epistasis, quantifying the relative importance of epistasis remains an open question. In this study we are able to identify 249 probes with at least one significant interaction given our experiment-wide threshold. Using the same threshold ($p < 2.91 \times 10^{-16}$) for the same data searching for only additive or dominance effects returns 517 probes out of the 7339 analysed with at least one significant effect, therefore the number of instances of epistasis are substantial. However, although many epistatic effects are observed, in terms of their contribution to complex traits a more important metric is the proportion of the variance that they explain. Ideally one would assess this question from a whole genome level but this is intractable for non-additive variance outside of in humans, but quantifying the ascertained effects detected in this study demonstrates that additive variance is overall a much larger component than epistatic variance. We observe that of the significant interactions the average additive contribution to the total phenotypic variance is 11.3\%, over three times higher than the epistatic contribution of 3.7\% (Figure \ref{fig:variancecomponents}). From an experiment-wide perspective, the proportion of the total phenotypic variance of all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold is 2.7\% (?????), whereas the epistasic signals detected in this study only explains 0.27\%. It must be stressed that this comparison strongly favours additive variance because the power of a 1 d.f. test exceeds that of an 8 d.f. test, and the non-additive variance at causal variants is expected to be grossly underestimated by observed SNPs in comparison to estimates for additive variance.

>>>>

\subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions, but our analysis shows this is unlikely to be sufficient. The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association (476 out of 549). Only 9 interactions were between SNPs that both had marginal effects and 64 were between SNPs that had no known marginal effects. Additionally, we observed that the largest epistatic variance component for the 529 interactions were proportionately divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms ($p = 0.22$ for divergence from expectation). This is expected as these traditional epistatic variance components are statistical formulations rather than biologically meaningful patterns.

We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common form of epistasis detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit a masking pattern such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect. Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets. Though the exact mechanism underlying this interaction requires further research, we observed that nine of the 14 \emph{trans} SNPs are located in intronic regions (significantly enriched, $p = 3.11 \times 10^{-9}$) and MBNL1 is involved in RNA modification and regulation of splicing, suggesting a putative mechanism by which epistatic interactions may regulate transcription.

\subsection{Contribution of epistatic variance}
Though we have discovered many instances of epistasis, quantifying the relative importance of epistasis remains an open question. In this study we are able to identify 249 probes with at least one significant interaction given our experiment-wide threshold. Using the same threshold ($p < 2.91 \times 10^{-16}$) for the same data searching for only additive or dominance effects returns 517 probes out of the 7339 analysed with at least one significant effect, therefore the number of instances of epistasis are substantial. However, although many epistatic effects are observed, in terms of their contribution to complex traits a more important metric is the proportion of the variance that they explain. Ideally one would assess this question from a whole genome level but this is intractable for non-additive variance outside of in humans, but quantifying the ascertained effects detected in this study demonstrates that additive variance is overall a much larger component than epistatic variance. We observe that of the significant interactions the average additive contribution to the total phenotypic variance is 11.3\%, over three times higher than the epistatic contribution of 3.7\% (Figure \ref{fig:variancecomponents}). From an experiment-wide perspective, the proportion of the total phenotypic variance of all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold is 1.9\%, whereas the epistatic signals detected in this study only explains 0.27\%. It must be stressed that this comparison strongly favours additive variance because the power of a 1 d.f. test exceeds that of an 8 d.f. test, and the non-additive variance at causal variants is expected to be grossly underestimated by observed SNPs in comparison to estimates for additive variance.
